---
title: "Master script for postfire analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Source functions, get data and plot

First we'll _source()_ (i.e. "run all code in") the scripts with the functions we made. Then we'll set the URL, read in the data with _download.NDVI()_, and plot it with _plot.NDVI()_.

```{r}
## Load required functions by running source() on the individual function files
if(file.exists("01_download.NDVI.R")) source("01_download.NDVI.R")
if(file.exists("02_plot.NDVI.R"))     source("02_plot.NDVI.R")
if(file.exists("03_negexp.R"))        source("03_negexp.R")
## Download NDVI data
URL = "https://raw.githubusercontent.com/jslingsby/BIO3019S_Ecoforecasting/master/data/modisdata.csv"
dat <- download.NDVI(URL)
# Convert "calendar_date" to postfire age in days since fire - assuming the first date in the times eries is the time of the fire 
dat$age <- (as.numeric(dat$calendar_date) - min(as.numeric(dat$calendar_date), na.rm = T))/365.25
## Plot overall NDVI time series
plot.NDVI(dat)
```

<br>

Q1: This plot suggests that Fynbos greenness (NDVI) as observed from satellite saturates with time since fire. Why do you think it saturates rather than increasing linearly with time?

>*Answer 1:*  
There is only a limited amount of land space available for plants to cover, therefore there is a hard limit to how much NDVI there can be. When the land available is fully covered by "greenness" there can be no further NDVI increase.  
Within the actual systems there are also limiting factors such as nutrient, sunlight, and water availability, competition, and herbivorous animals.  
These reasons explain why NDVI growth starts out rapidly and then slows down over time as both the available space starts becoming saturated, and the limiting factors make it increasingly more difficult for further growth to happen.
<br>

### 2. Fit models using Non-linear Least Squares (NLS)

Now we'll fit the simple and full negative exponential models using Non-linear Least Squares (NLS).

First the simpler model:

```{r}
## Simple model
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)
# fit model
fit_negexp <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda)),
                  data = dat, start = par, trace = F, 
                  control = nls.control(maxiter = 500))
# plot
plot.NDVI(dat = dat, fit = fit_negexp)
```

<br>

And let's look at the model summary with parameter estimates

```{r}
# print model summary
summary(fit_negexp)
```

<br>

Now the full model:

```{r}
## Full model
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)
# fit model
fit_negexpS <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda))
                   + A*sin(2*pi*age + (phi + pi/6*(3 - 1))), 
                   data = dat, start = par, trace = F, 
                   control = nls.control(maxiter = 500))
# plot
plot.NDVI(dat = dat, fit = fit_negexpS)
```


```{r}
# print model summary
summary(fit_negexpS)
```

<br>

Lots more parameters...

Q2: How do the estimates for the common parameters compare?

>*Answer 2:*  
All parameters were given the same starting values for the simple and the full model.  
The estimate for the "alpha" parameter decreased marginally by 0.04 in the full model. The estimate for parameter significance also did not change by much between the models.  
The estimate for the "gamma" parameter increased marginally by 0.04 in the full model. The estimate for parameter significance also did not change by much between the models.  
The estimate for the "lambda" parameter decreased noticeably by 0.19 in the full model. The estimate for parameter significance also changed greatly between models. With the t value for the parameter increasing from 5.500 (Pr(>|t|) = 1.84e-07) in the simple model to 7.846 (Pr(>|t|) = 1.25e-12) in the full model.
<br>

### 3. Compare NLS models using ANOVA

Modelers often want to know which of a set of models are better. One way to do this when comparing nested* models using least squares is using analysis of variance (ANOVA). In this case the `anova()` function will take the model objects as arguments, and return an ANOVA testing whether the full model results in a significant reduction in the residual sum of squares (and thus is better at capturing the data), returning an F-statistic, Degrees of Freedom (the difference in the number of parameters between the models) and p-value.

*i.e. one model is a subset of the other, as in our case

```{r}
anova(fit_negexp, fit_negexpS)
```

<br>

Q3: Which model is better?

>*Answer 3:*  
The full model is better than the simple model. This can be seen from the fact that the full model has a lower residual sum of squares, with a very large associated F-statistic of 39.207 (Pr(>F) = 4.12e-14).
<br>

Q4: How many degrees of freedom are there in this ANOVA and why (i.e. what are they)?

>*Answer 4:* 
This ANOVA has 2 numerator degrees of freedom which come from the fact that the simple model has 3 parameters and the full model has 5, resulting in a difference of 2 between the models.  
The simple model has 135 residual (denominator) degrees of freedom. Adding the 3 parameters onto that results in the simple model having 138 total degrees of freedom.  
The full model has 133 residual degrees of freedom. Adding the 5 parameters onto that results in the simple model having 138 total degrees of freedom.
<br>

### 4. Fit models using Maximum Likelihood Estimation (MLE)

First let's fit the simpler model:

```{r}
## Fit the simpler model using MLE
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)
# fit model
fit_negexpMLE <- fit.negexp.MLE(dat, par)
# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexp(fit_negexpMLE$par,dat$age), col = 'skyblue', lwd = 3)
```


```{r}
fit_negexpMLE
```

<br>

Then the full model:

```{r}
## Fit the full model using MLE
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)
# fit model
fit_negexpMLES <- fit.negexpS.MLE(dat, par)
# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexpS(fit_negexpMLES$par,dat$age), col = 'skyblue', lwd = 3)
```

```{r}
fit_negexpMLES
```

<br>

### 5. Compare MLE models using Akaike's information criterion (AIC)

Note that we can't compare our MLE models using ANOVA because our custom functions do not return full model fits like the `nls()` function - only the parameter estimates, negative log-likelihoods and a few other diagnostics.

Another way to compare models (and probably the most common) is using the Akaike information criterion (AIC), which is an estimator of prediction error (i.e. relative quality) of statistical models for a given set of data. 

The formula for the Akaike information criterion is:

$AIC = 2K -2(ln(L))$

Where:

- $k$ = the number of estimated parameters in the model
- $L$ = maximum value of the likelihood function for the model

Since we have our negative log likelihoods (i.e. $-ln(L)$ in the formula above), we can calculate the AICs and compare them.

```{r}
AIC_simple = 6 + 2*fit_negexpMLE$value
AIC_simple
AIC_full = 6 + 2*fit_negexpMLES$value
AIC_full
```

<br>

When comparing models, the lower the AIC the better, and in general a difference in AIC of 3 or more is analagous to the models being significantly different at an $\alpha$ of $p < 0.05$.

```{r}
AIC_simple - AIC_full
```

<br>

Q5: Is there a preferred model and if so, which one?

>*Answer 5:*  
The full model is preferred to the simple one since the full model has a much lower AIC than the simple model.
<br>

The nice thing about AIC is that the models you compare do not have to be nested like they do for ANOVA, as long as the data are the same. There are a few other constraints however... 

Here are the AIC scores for our pair of NLS models:

```{r}
AIC(fit_negexp, fit_negexpS)
```

<br>

You'll notice that these are completely different to the AICs for the MLE models...

Q6: Why is it not okay to compare the AIC of these NLS models with the AIC of the MLE models? Hint: type `?AIC` into the R console and do some reading.

>*Answer 6:*  
Although the AIC function in R can be used for many types of models, the theory of AIC itself requires that the log-likelihoods have been maximised.  
Therefore, even though we can get AIC values for the models fitted with non-linear least squares (i.e. not fitted with maximum likelihood estimation), it would not be applicable to use these AIC values in comparison to MLE fitted models, since they go against the base requirement of the AIC method and therefore could give unreliable results.
<br>